### 2017-01-20 conclusion.tex
Conlusion figure incorporated into text
Formatting updated
.svg figures incorporated into thesis correctly

\chapter{Conclusions and Future Work}
\label{sec:conclusion}

@@

@@

In Chapter \ref{sec:paper3}, both the occurrence and power of radar backscatter are used as proxies to indicate the ``amount'' of irregularities that exist in the ionosphere.  However, the trends from these quantities are rarely in good agreement, giving rise to the question what, if anything, does each of them indicate about the plasma structuring in the ionosphere.  The factors that control each need to be isolated, but this is nontrivial.  Echo occurrence generally agrees with the linear GDI theory predictions but echo power only does for relatively low power.  Irregularities may develop as predicted by linear GDI theory when their amplitudes are small but other nonlinear processes may impact larger density perturbations, such as shear-driven instabilities \citep{Gondarenko2006} or multistep processes \citep{Carlson2012}.  By the radar equation, the power returned should be proportional to the product of the background density and fractional density, squared, Equation \ref{eqn:fractional_density}.  However, there is no general consensus on what controls the fractional density in the \(F\) region, although it has been suggested that it is either constant or a function of convection speed, \(V_E\) \citep{Koustov1988,Haldoupis1990,Makarevich2014b}.  In Chapter \ref{sec:paper3}, we additionally propose it may be a function of the instability growth rate, \(\gamma\).  Because ground based radars are so often used to investigate ionospheric irregularities, this is an important issue to resolve to develop an accurate picture of how radar observations relate to plasma irregularities.

>>>>

\chapter{Conclusion}
\label{sec:conclusion}

@@

\begin{figure}
  \includegraphics[width=\textwidth]{conclusion.pdf}
  \caption[Irregularity production factors]{Summary of the main conclusions in this thesis as to what factors most effect pasma irregularity production.}
  \label{fig:conclusion}
\end{figure}

@@

In Chapter \ref{sec:paper3}, both the occurrence and power of radar backscatter are used as proxies to indicate the ``amount'' of irregularities that exist in the ionosphere.  However, the trends from these quantities are rarely in good agreement, giving rise to the question what, if anything, does each of them indicate about the plasma structuring in the ionosphere.  The factors that control each need to be isolated, but this is nontrivial.  Echo occurrence generally agrees with the linear GDI theory predictions but echo power only does for relatively low power.  Irregularities may develop as predicted by linear GDI theory when their amplitudes are small but other nonlinear processes may impact larger density perturbations, such as shear-driven instabilities \citep{Gondarenko2006} or multistep processes \citep{Carlson2012}.  By the radar equation, the power returned should be proportional to the product of the background density and fractional density, squared, Equation \ref{eqn:fractional_density}.  However, there is no general consensus on what controls the fractional density in the \(F\) region, although it has been suggested that it is either constant or a function of convection speed, \(V_E\) \citep{Kustov1988,Haldoupis1990,Makarevich2014b}.  In Chapter \ref{sec:paper3}, we additionally propose it may be a function of the instability growth rate, \(\gamma\).  Because ground based radars are so often used to investigate ionospheric irregularities, this is an important issue to resolve to develop an accurate picture of how radar observations relate to plasma irregularities.

### 2017-01-20 introduction.tex
.svg figures incorporated into thesis correctly

Small-scale structuring is generally considered to be all irregularities less than 100 m \citep{Kelley2009}.  Although small-scale irregularities are not responsible for scintillation that directly impact GPS/GNSS performance, they can be the easiest to study because of large data sets collected from high frequency ionospheric radars globally for several decades (discussed further in Section \ref{sec:superdarn}).  Because electrons move far more easily along magnetic field lines than across them, density perturbations tend to be aligned with the magnetic field and are often known as field-aligned irregularities (FAI), Figure \ref{fig:polarirreg}.  The irregularity classifications presented here are rough categories only and a high degree of coupling between different scales occurs, which makes plasma structures of all scales (large, intermediate, and small) important for plasma structuring processes.  For instance, large-scale structures such as polar patches may very well have internal intermediate- and/or small-scale structuring.  A turbulent cascade is often assumed to connect these different scales, but because various instruments or techniques usually only observe structuring on a particular scale, it is difficult to find direct evidence of this or how exactly it occurs \citep{Kinter1985,Tsunoda1985}.  A description of some of the particular instruments and models that are used in this thesis follows.

@@

Although the IRI model is very useful for providing background plasma density conditions where no instruments are available to make measurements, it is important to bear in mind that it is a model and therefor only represents the average conditions.  Although it is fairly accurate at mid-latitudes \citep{Coisson2006,Bilitza2012}, it does not necessarily represent variations within the polar cap well \citep{Themens2014,Makarevich2015}.  In particular, the height of the F-region peak tends to be underestimated in daytime and the bottomside thickness does not show realistic seasonal and diurnal variation \citep{Themens2014}.

@@

\subsection{Theory of Plasma Instabilities}
\label{sec:lit_instabilities}
Plasma structuring, especially at smaller scales, is often discussed in terms of the growth or damping of plasma waves, or instabilities.  As discussed previously in Section \ref{sec:ionosphere_regions}, the plasma characteristics change with altitude in the ionosphere, which causes a variety of different instability mechanisms to be operational depending on the region considered.  The Farley-Buneman instability (FBI), or the modified two-stream instability tends to be a major factor in the \(E\) region where ion inertial effects are strong \citep{Farley1963,Buneman1963}.  Also operational in the \(E\) region but more dominant at higher altitudes is the gradient-drift instability (GDI) \citep{Simon1963,Hoh1963,Linson1970}.  If a wave has a component along the magnetic filed line, the current-conductive instability (CCI) can be operational even if GDI is not \citep{Hoh1960,Ossakow1979,Chaturvedi1981}.  Additionally instability mechanisms emerge if shears are considered, when \(\nabla E \neq 0\), such as the Kelvin-Helmholtz instability (KHI) \citep{Kintner1977,DAngelo1965}.

@@

For the most part, the results of nonlinear numerical simulations very closely match expectations from linear GDI theory.  The spatial power spectra are anisotropic with most of the power concentrated in the direction where linear GDI growth is most favorable \citep{Keskinen1981,Keskinen1982,Gondarenko2001,Gondarenko2004b}.  However, very strong shears or increasing the impact of ion-inertial factors can cause isotropy \citep{Gondarenko2001,Gondarenko2006}.  The power spectra derived from these simulations are similar to those from experimental observations \citep{Baker1978,Kelley1979}, however it is computationally challenging to run simulations down to decameter scales and observations are generally limited to rocket and satellite techniques, which can only be used to calculate power spectra in one dimension \citep{Villain1986,Moen2012}.

>>>>

Small-scale structuring is generally considered to be all irregularities less than 100 m \citep{Kelley2009}.  Although small-scale irregularities are not responsible for scintillation that directly impact GPS/GNSS performance, they can be the easiest to study because of large data sets collected from high frequency ionospheric radars globally for several decades (discussed further in Section \ref{sec:superdarn}).  Because electrons move far more easily along magnetic field lines than across them, density perturbations tend to be aligned with the magnetic field and are often known as field-aligned irregularities (FAI), Figure \ref{fig:polarirreg}.  The irregularity classifications presented here are rough categories only and a high degree of coupling between different scales occurs, which makes plasma structures of all scales (large, intermediate, and small) important for plasma structuring processes.  For instance, large-scale structures such as polar patches may very well have internal intermediate- and/or small-scale structuring.  A turbulent cascade is often assumed to connect these different scales, but because various instruments or techniques usually only observe structuring on a particular scale, it is difficult to find direct evidence of this or how exactly it occurs \citep{Kintner1985,Tsunoda1985}.  A description of some of the particular instruments and models that are used in this thesis follows.

@@

Although the IRI model is very useful for providing background plasma density conditions where no instruments are available to make measurements, it is important to bear in mind that it is a model and therefor only represents the average conditions.  Although it is fairly accurate at mid-latitudes \citep{Coisson2006,Bilitza2012}, it does not necessarily represent variations within the polar cap well \citep{Themens2014,Makarevich2015b}.  In particular, the height of the F-region peak tends to be underestimated in daytime and the bottomside thickness does not show realistic seasonal and diurnal variation \citep{Themens2014}.

@@

\subsection{Theory of Plasma Instabilities}
\label{sec:lit_instabilities}
Plasma structuring, especially at smaller scales, is often discussed in terms of the growth or damping of plasma waves, or instabilities.  As discussed previously in Section \ref{sec:ionosphere_regions}, the plasma characteristics change with altitude in the ionosphere, which causes a variety of different instability mechanisms to be operational depending on the region considered.  The Farley-Buneman instability (FBI), or the modified two-stream instability tends to be a major factor in the \(E\) region where ion inertial effects are strong \citep{Farley1963,Buneman1963}.  Also operational in the \(E\) region but more dominant at higher altitudes is the gradient-drift instability (GDI) \citep{Simon1963,Hoh1963,Linson1970}.  If a wave has a component along the magnetic field line, the current-conductive instability (CCI) can be operational even if GDI is not \citep{Hoh1960,Ossakow1979,Chaturvedi1981}.  Additionally instability mechanisms emerge if shears are considered, when \(\nabla E \neq 0\), such as the Kelvin-Helmholtz instability (KHI) \citep{Kintner1977,DAngelo1965}.

@@

For the most part, the results of nonlinear numerical simulations very closely match expectations from linear GDI theory.  The spatial power spectra are anisotropic with most of the power concentrated in the direction where linear GDI growth is most favorable \citep{Keskinen1981a,Keskinen1981b,Keskinen1982a,Gondarenko2001,Gondarenko2004b}.  However, very strong shears or increasing the impact of ion-inertial factors can cause isotropy \citep{Gondarenko2001,Gondarenko2006}.  The power spectra derived from these simulations are similar to those from experimental observations \citep{Baker1978,Kelley1979}, however it is computationally challenging to run simulations down to decameter scales and observations are generally limited to rocket and satellite techniques, which can only be used to calculate power spectra in one dimension \citep{Villain1986,Moen2012}.

### 2017-01-20 paper1.tex
.svg figures incorporated into thesis correctly

Production of these irregularities is reasonably well understood on a microphysical level, with GDI promoting irregularity growth for certain orientations of the density gradient vector relative to the convection electric field \citep{Kesknien1982a,Kesknien1982b,Makarevich2014c}. The role of global factors, on the other hand, such as geomagnetic activity or solar wind conditions is much less investigated. In particular, the extent of the solar control is largely unknown, both in terms of its direct effects due to varying ionization and  indirect effects due to varying interplanetary magnetic field (IMF) as explained below. This is particularly important to processes occurring in the polar cap where some of the strongest differences in solar illumination conditions occur and where geomagnetic field lines are directly connected to the IMF.

>>>>

Production of these irregularities is reasonably well understood on a microphysical level, with GDI promoting irregularity growth for certain orientations of the density gradient vector relative to the convection electric field \citep{Keskinen1982a,Keskinen1982b,Makarevich2014c}. The role of global factors, on the other hand, such as geomagnetic activity or solar wind conditions is much less investigated. In particular, the extent of the solar control is largely unknown, both in terms of its direct effects due to varying ionization and  indirect effects due to varying interplanetary magnetic field (IMF) as explained below. This is particularly important to processes occurring in the polar cap where some of the strongest differences in solar illumination conditions occur and where geomagnetic field lines are directly connected to the IMF.

### 2017-01-20 paper2.tex
.svg figures incorporated into thesis correctly

Figure \ref{fig:1}c presented the growth rate pattern for one particular orientation of the gradient reversal and plasma convection. Figure \ref{fig:2} presents the same analysis for four different orientations and locations of the gradient reversal. In all cases, the convection direction is assumed to be perpendicular to the gradient reversal direction (\(\phi-\theta=90\deg\)).  This is because large-scale structures, such as polar patches tend to drift perpendicular to their elongation direction \citep{Makarevich2004c,Makarevich2015b}. This requirement was convenient for presenting different growth rate patterns, but will be relaxed later in this study.

@@

Although this study is relevant to observations with obliquely scanning HF radars, care should be taken with how the results are used to interpret experimental data.  In this study, only the linear growth rate of GDI was considered. Even though radar backscatter power and occurrence trends are often interpreted using the GDI growth rates in the linear regime \citep[e.g.][]{Oksavik2012,Moen2012}, there are other factors that may affect echo power and occurrence.  In the polar cap, HF radars rely on beam refraction through a dense ionosphere to observe field-aligned irregularities \citep{Bristow2011,Koustov2014}.  In addition, high background electron density in both the E and F regions can smooth gradients required for GDI and limit instability growth \citep{Ruohoniemi1997,Koustov2004}.  Large conductivity in the E region can also short out F region irregularities \citep{Danskin2002,Kane2012,Lamarche2015}.  An enhanced D region may act to attenuate backscatter signals, although this effect is not always obvious \citep{Dansin2002}.  Nonlinear processes in the F region have been theoretically shown to alter the dependence on the propagation direction while still showing significant anisotropy similar to that predicted by the linear theory \citep{Keskinen81,Keskinen82a,Keskinen83}.  The situation in the E region appears to be somewhat different, with some echoes also exhibiting strong anisotropy \citep[e.g.][]{Ierkic1980,Makarevich2008} and other echoes showing isotropic power variation \citep{Ierkic1980,Koustov2001} presumably because of nonlinear processes.  The results of the present study are thus more likely to apply in the F region.

>>>>

Figure \ref{fig:1}c presented the growth rate pattern for one particular orientation of the gradient reversal and plasma convection. Figure \ref{fig:2} presents the same analysis for four different orientations and locations of the gradient reversal. In all cases, the convection direction is assumed to be perpendicular to the gradient reversal direction (\(\phi-\theta=90\deg\)).  This is because large-scale structures, such as polar patches tend to drift perpendicular to their elongation direction \citep{Makarevitch2004c,Makarevich2015b}. This requirement was convenient for presenting different growth rate patterns, but will be relaxed later in this study.

@@

Although this study is relevant to observations with obliquely scanning HF radars, care should be taken with how the results are used to interpret experimental data.  In this study, only the linear growth rate of GDI was considered. Even though radar backscatter power and occurrence trends are often interpreted using the GDI growth rates in the linear regime \citep[e.g.][]{Oksavik2012,Moen2012}, there are other factors that may affect echo power and occurrence.  In the polar cap, HF radars rely on beam refraction through a dense ionosphere to observe field-aligned irregularities \citep{Bristow2011,Koustov2014}.  In addition, high background electron density in both the E and F regions can smooth gradients required for GDI and limit instability growth \citep{Ruohoniemi1997,Koustov2004}.  Large conductivity in the E region can also short out F region irregularities \citep{Danskin2002,Kane2012,Lamarche2015}.  An enhanced D region may act to attenuate backscatter signals, although this effect is not always obvious \citep{Danskin2002}.  Nonlinear processes in the F region have been theoretically shown to alter the dependence on the propagation direction while still showing significant anisotropy similar to that predicted by the linear theory \citep{Keskinen1981b,Keskinen1982a,Keskinen1983}.  The situation in the E region appears to be somewhat different, with some echoes also exhibiting strong anisotropy \citep[e.g.][]{Ierkic1980,Makarevich2008} and other echoes showing isotropic power variation \citep{Ierkic1980,Koustov2001} presumably because of nonlinear processes.  The results of the present study are thus more likely to apply in the F region.
